**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-05-06 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Wroking on last section

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap = "Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions."}
# linear response, correlated errors (subject effect)
par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

@@

```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap="Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for time and original datapoints. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points."}

@@

For the data in Figure \@ref(fig:sim-smooth-plot),B it is of interest to know the effect that each treatment has in oxygen saturation. In a practical sense, this implies an interaction model where smooths are created for _Group_ and _Day_, respectively. The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term. Using the package `mgcv` The model then becomes:

`m1<-gam(StO2~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`

This syntax specifies that the object `m1` will store the model, and that the change in $StO_2$ is modeled by independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),C).

@@

m_1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)

@@

#adds the predictions to the grid and creates a confidence interval
f_predict<-f_predict%>%
    mutate(fit = predict(m_1,f_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(m_1, f_predict,se.fit = TRUE,type='response')$se.fit)

@@

>>>>

(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(l-q-response-caption)' }
# linear response, correlated errors (subject effect)
par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

@@

(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.

```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}

@@

For the data in Figure \@ref(fig:sim-smooth-plot),B it is of interest to know the effect that each treatment has in oxygen saturation. In a practical sense, this implies an interaction model where smooths are created for _Group_ and _Day_, respectively. The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using $\textsf{R}$ and the package `mgcv` the model syntax is:

`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`

This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),C). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia]. A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)

@@

#adds the predictions to the grid and creates a confidence interval
f_predict<-f_predict%>%
    mutate(fit = predict(m1,f_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(m1, f_predict,se.fit = TRUE,type='response')$se.fit)

@@

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data that has missing observations. Again, using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and a GAM model with interaction can be fitted to the remaining observations.

```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
#missing data
#create a sequence of 40 random numbers between 1 and 100, these numbers will
#correspond to the row numbers to be randomly erased from the original dataset
missing <- sample(1:100, 40)

#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA

ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))

#create a new dataframe, remove the StO2 column
dat_missing <- dat_sim[,-1]

#add NAs at the ind positions
dat_missing$StO2_sim[ind]<-NA

#Count the number of remaining observations per day (original dataset had 10 per group per day)
dat_missing %>%
    group_by(Day,Group) %>%
    filter(!is.na(StO2_sim))%>%
  count(Day)

#plot the simulated data
f4<-ggplot(dat_missing, aes(x = Day, y = StO2_sim, color = Group)) +
    geom_point() +
  stat_summary(aes(y = StO2_sim,
                     group=Group), 
                 fun=mean, geom="line",size=1)+
    ggtitle("Missing data") +
  labs(y='StO2')+
  scale_color_aaas()+
    theme_classic()

f2+f4+plot_layout(guides = 'collect') &
  ylim(c(0,75)) &
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )&
  scale_color_aaas()

#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
#appraise the model
appraise(mod_m1)

m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#adds the predictions to the grid and creates a confidence interval
m_predict<-m_predict%>%
    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)

ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    facet_wrap(~Group)+
    geom_point(colour='black',size=1,alpha=0.5)+
    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
                data=m_predict, alpha=0.5,
                inherit.aes=FALSE) +
    geom_line(aes(y=(fit),color=factor(Group)),
              size=1,data=m_predict,show.legend = FALSE)

comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp2))

missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', 
         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
                              )))+
  scale_x_continuous(breaks=c(0,2,5,7,10))+
  theme_classic()+
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )
```
